---
title: "Third language acquisition and individual differences"
subtitle: 'A quantitative case for case studies'
author  : "Kyle Parrish"
date    : "</br>Winter 2024</br>Last update: `r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: [default, hygge, my-css.css, rutgers-fonts]
    lib_dir: libs
    nature:
      countIncrementalSlides: false
      ratio: 16:9
---

```{r, setup, include=FALSE}
library("fontawesome")
library("RefManageR")
knitr::opts_chunk$set(
  comment = '', fig.width = 6, fig.height = 6, message = FALSE, warning = FALSE, echo = FALSE
)
```

```{r xaringan-themer, include=FALSE}
library(xaringanthemer)
style_duo_accent(
  # primary_color = "#345865",
  primary_color = "#B56B6F",
  secondary_color = "#2c8475",
  black_color = "#4242424",
  white_color = "#FFF",
  base_font_size = "25px",
  # text_font_family = "Jost",
  # text_font_url = "https://indestructibletype.com/fonts/Jost.css",
  header_font_google = google_font("Times New Roman", "400", "700"),
  header_font_weight = "400",
  inverse_header_color = "#eaeaea",
  title_slide_text_color = "black",
  text_slide_number_color = "#9a9a9a",
  text_bold_color = "#f79334",
  code_inline_color = "#B56B6F",
  code_highlight_color = "transparent",
  link_color = "#2c8475",
  table_row_even_background_color = lighten_color("#345865", 0.9),
  extra_fonts = list(
    google_font("Times New Roman", "400")
  ),
  colors = c(
    green = "#31b09e",
    "green-dark" = "#2c8475",
    highlight = "#87f9bb",
    purple = "#887ba3",
    pink = "#B56B6F",
    orange = "#f79334",
    red = "#dc322f",
    `blue-dark` = "#002b36",
    `text-dark` = "#202020",
    `text-darkish` = "#424242",
    `text-mild` = "#606060",
    `text-light` = "#9a9a9a",
    `text-lightest` = "#eaeaea"
  ),
  extra_css = list(
    ".remark-slide-content h3" = list(
      "margin-bottom" = 0, 
      "margin-top" = 0
    ),
    ".smallish, .smallish .remark-code-line" = list(`font-size` = "0.9em")
  )
)
xaringanExtra::use_xaringan_extra(c("tile_view", "animate_css", "tachyons", "share_again"))
xaringanExtra::use_extra_styles()
```


```{r components, include=FALSE}
slides_from_images <- function(
  path,
  regexp = NULL,
  class = "hide-count",
  background_size = "contain",
  background_position = "top left"
) {
  if (isTRUE(getOption("slide_image_placeholder", FALSE))) {
    return(glue::glue("Slides to be generated from [{path}]({path})"))
  }
  if (fs::is_dir(path)) {
    imgs <- fs::dir_ls(path, regexp = regexp, type = "file", recurse = FALSE)
  } else if (all(fs::is_file(path) && fs::file_exists(path))) {
    imgs <- path
  } else {
    stop("path must be a directory or a vector of images")
  }
  imgs <- fs::path_rel(imgs, ".")
  breaks <- rep("\n---\n", length(imgs))
  breaks[length(breaks)] <- ""

  txt <- glue::glue("
  class: {class}
  background-image: url('{imgs}')
  background-size: {background_size}
  background-position: {background_position}
  {breaks}
  ")

  paste(txt, sep = "", collapse = "")
}
options("slide_image_placeholder" = FALSE)
```


```{r}
library(readr)
library(here)
library(sjPlot)
mod0 = read_rds(here("practice_talk", "data", "study1", "mod0.rds"))
```


## Third Language Learning is **not** Second Language Learning 

--

.pull-left[
**How are they different?**
- .blue[L2 learning] in adulthood is largely predicted by **L1-L2 differences**

- .red[L3 learning] is distinct because there is evidence **both the L1 and L2 can play a role**

- We do not know **which factors** modulate relative access to the L1 or L2 in L3 speakers.
]

--

.pull-right[
**Why is it important to research L3 learning?**

- Bilingualism is becoming the norm worldwide: 50%+ bilingual population.

- L3 learning is common at Texas Tech - Languages other than Spanish and English at Texas Tech are likely L3s: 27% Hispanic enrollment.

- No standards for L3 development exist.
]

---

## How do the L1 and the L2 play a role in L3 learning?

Conflicting evidence has been found:

--

- **Typological similarity** of the languages overall results in L1 or L2 transfer.
 - Spanish and Portuguese are more similar to each other than English

--

- **Order of acquisition** biases influence despite similarity.
 - Late-learned L2 influence is different from L1 influence independent of similarity (L2 status effect).
 
--

- **Structural similarity** of specific parts of the languages results in L1 or L2 influence on a property-by-property basis. 
 - Adjective order is different in Romance Languages than English, but the position of verbs is largely the same. 

--


---

class: title-slide-section, middle
<br>
<br>
<br>
<br>
## Conflicting narrative conclusions in L3 work are based on **misunderstandings** of statistical tools 

<!---
Importance of figuring this out 

Narrative conclusions are largely the basis for models of third language acquisition. 

Models predict behavior and are some of the few points of reference in interventions for L3 learning. 

--->

---

## The Typological Primary Model (TPM)

--

A major model in L3 acquisition, it predicts that overall psychotypological similarity predicts **whole language influence** during the initial stages of L3 development, despite order of acquisition.

--

.pull-left[
**Mirror Image Design**

.blue[L1 Spanish] - .red[L2 English] 

.red[L1 English] - .blue[L2 Spanish]


These groups should behave similarly learning .green[L3 Portuguese] 
]

--

.pull-right[
Practical implications would be that we should treat **L3 learners the same as L2 learners** if we can determine their transfer source language. 
]

---

## The evidence for the TPM is based on a statistical misunderstanding

Many studies have run tests on these so-called mirror image groups and found no significant differences.

--

**Misunderstanding 1**: Statistical Insignificance is not practical equivalence 

---

These studies are done on groups and often have low sample sizes 

--

Low sample studies run high risks of **false negative findings**, especially when the true effect size decreases - big effects require less data to find due to sampling error 

---

The original study in which the model was proposed, the primary evidence was a null effect between two groups in which one had 12 and the other had 15 participants 

Although it is possible that the true effect was 0, the original work did not generate enough data to say so 

Put better, In terms of differences, only enough data was present to detect a large effect. 

Thus, the results are better understood as inconclusive, in which the true effect between groups could be 0, a small effect, or a medium effect. 

We can be more certain the effect is 0, and not small or medium, by gathering more data. 

---

So I replicated it! Link to study 

---

Power graph 

--

The graph shows that 107 per group was needed to detect equivalence between groups, given a specific effect size as eq bounds 

---

The study:
XXX L3 speakers of P 
XX L1 Sp
XX L1 Eng
Looking for differences performance on the tasks between the groups 

---

Tasks

---

Results 

---

The results show that, out of 4 comparisons, 2 replicate, 1 does not, and 1 is still inconclusive 

Evidence for full transfer is almost non-existent at the group level when we consider statistical power 

---

Now, let’s look at intermediate evidence 

--

Misunderstanding 2: Multilevel models do not predict performance for most participants when there are latent variables in the data

<!-- This time I’ll use my own work to zoom in on, and my own assumptions -->

---

L3 phonetics has several published papers focusing on VOT (voice onset time)

---

VOT is…. definition 

---

Show study 2 

---

The findings were taken as evidence that the L3 fell between L1 and L2 - and L3 VOT > L2 VOT - L2 influence was occurring 

---

But, during data segmentation, I noticed trends - many participants were very English like, and many were Spanish like, and least of all intermediate 

---

So the question becomes…what if the group model simply shows us a central tendency, and does not actually predict the behavior of most participants?

--

Simple example - 6 people have scores  25, 25, 50, 50, 75, 75

Mean is X but does not apply to 66% of participants  

---

I wanted to expand this to an L3 design and quantify the implications - how likely is the body of literature to be biased? In other words... how many data points result in the bias at the group level? 

---

Show study 3 

---

# Take-aways 

The results show that group models are biased to show intermediate group results even when it is only true for some people 

Only when we have ~35 sounds per language do we accurately sub-categorize them

---

# Implicaitons 

--

This is important for language teaching because predictions in learning are made on the basis of known language-learned language similarities and contrasts 

--

But how do we generalize the information we glean from case studies? 

--

Long-term, commonalities with high statistical power and more certainty arise.

Latent variables are identified and inspire specific group studies 

More nuanced information can be gathered 

New techniques can be used - Random forests 

--

# A Path forward at Texas Tech 

--

**Lubbock and Texas Tech are ideal places to study L3**

Robust cross linguistic evidence - X major and X minors

High percentage of bilinguals (27%)

--

**Uncovering Latent variables**

K-12 World Language teachers conduct “experiments” all the time 

Establishing a bridge between research and practice can be mutually beneficial 














---
# Overview of the Talk

--

.large[
.blue[Part 1]: Background and context.
]

--

.large[
.blue[Part 2]: **False negatives** are highly probable when participants are low.
]

--

.large[
.blue[Part 3]: **Group-level effects** of models may **not apply to a majority of participants** in a data set.
]

--

.large[
.blue[Part 4]: Work at **Texas Tech**: Many Ln case studies and corroboration within the individual.
]

--

.large[
.blue[Part 5]: Wrap-up and Q & A.
]
---

class: title-slide-section, middle
<br>
<br>
<br>
<br>
## Introduction

---

class: title-slide-section, middle
<br>
<br>
<br>
<br>
## **Third Language (L3) acquisition** is not simply learning another L2.

--

### A body of work has shown cases of L1 and L2 influence.

---

# Second and Third Language Research

--

.large[
- Third language learners likely **represent at least 30%** of language learners.
]

--

.large[
- There are **not standards for L3 developmental paths**, and how they are influenced by their previously known linguistic knowledge, especially Heritage Speakers. 
]

--

.large[
- Perceptions of **typological similarity** between languages, **order of acquisition** of previously acquired languages and **proficiency and dominance** all have impacts on L3 learning
]

--

.large[
- ...**NOT** just the L1!
]


---

# Psychotypological Proximity and the L3

--

.big[
**The Typological Primacy Model** (Rothman, 2011; 2013; 2015).
]

--

.large[
- Either the L1 or the L2 transfers **wholesale** and serves as the intial state of the L3.
- The choice occurs during the first ~30 hours of instruction.
- Between the L1 and the L2, the one that is perceived as closer to the L3 transfers. 
]

--

*visual of L1 eng L2 span l3 BP*

---

# Linguistic Proximity and the L3

--

.big[
**The Linguistic Proxmity Model** (Westergaard et al., 2017).
]

--

.large[
- Both the L1 or the L2 can transfer on a **property-by-property** basis.
- Transfer is incremental and allows for both facilitative and non-facilitative effects.
- Learning occurs by parsing input and concerns underlying structural representations, not psychotypological proximity.
]

--

*Intermediate performace from Westergaard et al 2017*

---

# Statistical Approaches to L3 studies 

.large[
L3 research has been carried out using **small sample experimental groups**, and has produced conflicting findings. 
]

--

.pull-left[
**Null effects** as evidence for equivalence. 

 - Many studies provide support for the TPM on the basis of a non-significant finding on a Null Hypothesis Significance Test (NHST).
 - The number of participants per group is generally low in these studies (< 15).
]

--

.pull-right[
**Mixed effects models** as evidence for intermediate group performance.

- Studies, particularly in L3 phonetics, use mixed effects models to predict a linguistic outcome as a function of language (L1, L2 or L3).
- Findings of the models have shown the group effect that support the LPM (e.g., L1 < L3 < L2)
]

--

.large[
Both of these approaches have concerning risks - the field should move toward **multi-case studies** and focus on the individual, not the group. 
]

---

class: title-slide-section, middle
<br>
<br>
<br>
<br>
## Part 2: **False negatives** are more probable as the quantity of participants decreases.

--

### Statistical Insignificance is not wholesale transfer in L3 Acquisition: an approximate replication of Rothman (2011)

.footnote[[Link to article](DOI:10.1017/S0272263124000342)]

---

# Motivation for Replication 

.large[
Low sample studies are associated with high false negative rates.
]

--

.large[
The original study made its claim on the basis of a a null (negative effect) in two small sample groups in (n = 12 and n = 15).
]

--

.large[
This is not enough data to provide evidence for equivalence, and for differences can only detect a very large effect (Cohen's d > 1) reliably. 
]

--

.large[
The replication examined the hypothesis that the original finding was a false negative caused by low power.
]

---

# Determiner Phrase Syntax in Spanish, Portugese and English 

--

.large[
In Spanish and Portuguese, but not English, adjectives can occur in a pre or post nominal position, and this impacts its meaning.
]

--

.large[
Example: inc screenshot 
]

---


.large[
The studies used two tasks to examine how noun-adjective order within a DP impacted meaning.

Context-based collocation Task 

Semantic Interpretation Task
]


---

# Participants 

**211** participants total split into two groups 

**96** L1 English L2 Spanish L3 Portuguese 

**115** L1 Spanish L2 English L3 Portuguese 

---

# Research Questions and Hypotheses 

RQ1: Will the Spanish L1 and English L1 groups perform the same in perception and production of adjective-noun order in determiner phrases in their L3?

RQ2: Will the lack of a main effect in the models also be practically equivalent when the equivalence bounds are d = ± .4?

---

# Analysis 

We have four comprisons to look at:

2 tasks x 2 positions between groups.






---

# Results 



---