---
title: "Individual differences must underlie L3 acquisition"
subtitle: 'Job Talk'
author  : "Kyle Parrish"
date    : "</br>Winter 2024</br>Last update: `r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: [default, hygge, my-css.css, rutgers-fonts]
    lib_dir: libs
    nature:
      countIncrementalSlides: false
      ratio: 16:9
---

```{r, setup, include=FALSE}
library("fontawesome")
library("RefManageR")
knitr::opts_chunk$set(
  comment = '', fig.width = 6, fig.height = 6, message = FALSE, warning = FALSE, echo = FALSE
)
```

```{r xaringan-themer, include=FALSE}
library(xaringanthemer)
style_duo_accent(
  # primary_color = "#345865",
  primary_color = "#B56B6F",
  secondary_color = "#2c8475",
  black_color = "#4242424",
  white_color = "#FFF",
  base_font_size = "25px",
  # text_font_family = "Jost",
  # text_font_url = "https://indestructibletype.com/fonts/Jost.css",
  header_font_google = google_font("Times New Roman", "400", "700"),
  header_font_weight = "400",
  inverse_header_color = "#eaeaea",
  title_slide_text_color = "black",
  text_slide_number_color = "#9a9a9a",
  text_bold_color = "#f79334",
  code_inline_color = "#B56B6F",
  code_highlight_color = "transparent",
  link_color = "#2c8475",
  table_row_even_background_color = lighten_color("#345865", 0.9),
  extra_fonts = list(
    google_font("Times New Roman", "400")
  ),
  colors = c(
    green = "#31b09e",
    "green-dark" = "#2c8475",
    highlight = "#87f9bb",
    purple = "#887ba3",
    pink = "#B56B6F",
    orange = "#f79334",
    red = "#dc322f",
    `blue-dark` = "#002b36",
    `text-dark` = "#202020",
    `text-darkish` = "#424242",
    `text-mild` = "#606060",
    `text-light` = "#9a9a9a",
    `text-lightest` = "#eaeaea"
  ),
  extra_css = list(
    ".remark-slide-content h3" = list(
      "margin-bottom" = 0, 
      "margin-top" = 0
    ),
    ".smallish, .smallish .remark-code-line" = list(`font-size` = "0.9em")
  )
)
xaringanExtra::use_xaringan_extra(c("tile_view", "animate_css", "tachyons", "share_again"))
xaringanExtra::use_extra_styles()
```


```{r components, include=FALSE}
slides_from_images <- function(
  path,
  regexp = NULL,
  class = "hide-count",
  background_size = "contain",
  background_position = "top left"
) {
  if (isTRUE(getOption("slide_image_placeholder", FALSE))) {
    return(glue::glue("Slides to be generated from [{path}]({path})"))
  }
  if (fs::is_dir(path)) {
    imgs <- fs::dir_ls(path, regexp = regexp, type = "file", recurse = FALSE)
  } else if (all(fs::is_file(path) && fs::file_exists(path))) {
    imgs <- path
  } else {
    stop("path must be a directory or a vector of images")
  }
  imgs <- fs::path_rel(imgs, ".")
  breaks <- rep("\n---\n", length(imgs))
  breaks[length(breaks)] <- ""

  txt <- glue::glue("
  class: {class}
  background-image: url('{imgs}')
  background-size: {background_size}
  background-position: {background_position}
  {breaks}
  ")

  paste(txt, sep = "", collapse = "")
}
options("slide_image_placeholder" = FALSE)
```


```{r}
library(readr)
library(here)
library(sjPlot)
mod0 = read_rds(here("practice_talk", "data", "study1", "mod0.rds"))
```

class: title-slide-section, middle
<br>
<br>
<br>
<br>
## Knowledge of many languages is a **bridge to greater oppotunities** that have a lasting positive impact on the individual and society 

--

### Social and class mobility 

--

### Greater cultural understanding 

--

### Employment opportunities 

---


class: title-slide-section, middle
<br>
<br>
<br>
<br>
## This is especially true in Lubbock 

--

XX% Hisp 
XX% L2 

--

XX class X Y Z 

---


class: title-slide-section, middle
<br>
<br>
<br>
<br>
## But there is a problem: in adulthood, learning a new language is a difficult undertaking.

--

### Research shows that there is a wide range of ultimate attainment. 

---

## The big question: 

--

### **What factors** play a role in ultimate attainment, causing this variation? 

--

.large[
.pull-left[


Can we **predict** a person's ultimate attainment in a new language?
]]

--

.large[
.pull-right[
Can we **intervene** in the process and optimize language learning ability? 
]]

---


class: title-slide-section, middle
<br>
<br>
<br>
<br>
## Contextualization 

---

# Frequent assumptions in L2 acquisition 

.pull-left[
1. During L2 learning, the L1 plays an important role.

2. Third language are subsumed under L2 learning (guided by L1-L2 differences)
]

---

# The evidence suggests L3 learning and L2 learning are different 

Research in L3 learners has shown that they are not simply learning a second L2 - both the L1 and L2 play a role.

--

Findings conflict on exactly how: 
 - no access to L1/L2
 - access to both
 - only access to helpful aspects

--

## So - it is unclear given the mixed findings whether L3 acquisition follows any single one of these patterns, or, alternatively, if they could **all explain the process for certain individuals**, explained by particular differences. 

--

If we better understand what individual differences are relevant, we can leverage our work to make a more socially responsible impact that can be utilized by its beneficiaries

---

# A relevant example 

A lot of evidence comes from L3 speakers with a **monolingual upbringing** and **does not consider Heritage bilinguals** learning a new language.

The problem here is that evidence to inform language acquisition based solely on learners brought up monolingual **would not be an equitable investment** in the population of L3 learners and disproportionately benefit only some learners. 

Differences in Heritage L3 learners and L3 monolingual upbringing learners: 
- more balanced bilinguals are impacted by their more dominant languages
- monolingual upbringing sensitive to other factors.

There are still many conflicting findings to sort out. 

---

class: title-slide-section, middle
<br>
<br>
<br>
<br>
## One way to sort out the mixed findings is **replication**. If a general finding is true, we should generate similar data if we carry out the same experiment on the same population.

---

TPM replication study 

Partial replication - probably because of an issue with power and narratvie conclustions on the basis of the statistical tests 

---

The study did DP syntax and looked at L3 BP or Italian by two bilingaul groups. (examples in the languages)

---

The groups were L1 English L2 Spanish and L1 Italian L2 English

The question was whether these two would treat the DP strucutres differently, if not, the assumption was that they were both influenced by Sp, if so, order of ac plays a role.

---

The study concluded that Spanish, whether it was the L1 or L2, impcated the interpretation of DP sentences 

---

The issue was that few participants were included in the study - brining in the chance that power was low. 

--

Low power would increase the risk of a false negative finding.

---

I replicated this study, creating new materials and keeping the L3 constant 

---

The results revealed a differences as the sample increased, while also suggesting that the original study may have been a false negative, given that it could only relaibly detect rather large effects. 

---


class: title-slide-section, middle
<br>
<br>
<br>
<br>
## Additionally, if these findings reflect language-wide influence or a universal language learning process, **they should generalize**.

--

Many linguistic structures or levels within the individual.

--

New language triplets. 

---

## So, the big questions are, within L3 findings - 

--

What replicates?

--

What generalizes among languages? 

--

What differs as a function of the individual?

---

# Study 1: Replication 

TPM stuff

Partial replication.... but no evidence on the TOST. 

---

# Study 2: Replication



# The next wave of questions:

What individual factors are associated with the overcoming of non-facilitative influence from the L1 or L2? Which factors impede access to otherwise helpful influence?

Are all linguistic levels subject to the same influence?

Will online data provide new insights? Are there important differences in how CLI manifests depending upon processing load?

Do findings generalize to new language triplets/L3s?
